Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-05 index.qmd
section:issues; process:code

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - probability
  - Thurstone
  - comparative judgement
  - directed acyclic graph
  - structural causal models
  - statistical modeling
abstract: |
  (to do)
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

@@

<!-- 1. What does CJ do? -->
[In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level. After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest. This method has been successfully employed in assessing the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].]{style="color:red;"}

@@

<!-- 2. Where are we now with DCJ? -->
Numerous studies have documented the effectiveness of CJ in assessing various traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Additionally, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023], while research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.

@@

<!-- 2. But there is no structural model, only post estimation analyses -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing. The literature indicates that the scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
However, the statistical literature cautions against using estimated scores to conduct additional analyses or tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].

@@

@@

<!--   From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008.  -->

<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->

<!--   Table with assumptions -->

@@

@@

## References {.unnumbered}

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - Probability
  - Directed Acyclic Graphs
  - Bayesian methods
  - Thurstonian model
  - Comparative judgement
  - Structural Causal Models
  - Statistical modeling
abstract: |
  (to do)
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

```{r}
#| results: hide
#| label: code-package

# # load packages
# libraries = c('RColorBrewer','stringr','dplyr','tidyverse',
#               'rechape2','knitr','kableExtra',
#               'rstan','StanHeaders','runjags','rethinking')
# sapply(libraries, require, character.only=T)

```

```{r}
#| results: hide
#| label: code-usd

# # load functions
# main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript'
# source( file.path( main_dir, 'code', 'user-defined-functions.R') )

```

@@

<!-- 1. What does CJ do? -->
In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level. After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest. This method has been successfully employed in assessing the quality of written texts, where quality describes the underlying trait of interest and the texts serve as the stimuli [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. Where are we now with DCJ? -->
Numerous studies have documented the effectiveness of CJ in assessing various traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Additionally, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of CJ studies, the literature's unsystematic and fragmented research approaches have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.

@@

<!-- 2. But there is no structural model, only post estimation analyses -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing. The scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
However, the statistical literature cautions against using estimated scores to conduct additional analyses or tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020]. A type I error results when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].

@@

<!-- 1. What are the five cases of Thurstone? -->

@@

From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008.

  What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution

  Table with assumptions

@@

## Appendix A: Ignoring uncertainty {#sec-appendix-A}

```{r}
#| label: code-AppA-simulation

# # simulate units and sub-units
# USd = sim_units_trait( Un=10,
#                        Us=c(1,1),
#                        Ub=c(0,-0.2,0.2),
#                        Useed = 45789,
#                        Sn=6,
#                        Ss=c(0.8,0.8),
#                        Sb=c(0,0,0),
#                        Sseed = 9478 )
# 
# # simulate judges bias
# Jd = sim_judges_bias( Jn=10,
#                       Js=c(0.02,0.02),
#                       Jseed = 79985 )
# 
# 
# # simulate full comparison data
# d = cm_full( USd=USd$USd,
#              Jd=Jd$Jd )
# 
# 
# # set data in list
# dL = list_data( d )
# str(dL)

```

## Appendix B: The five cases of Thurstone {#sec-appendix-B}

<!-- Thurstone's comparative judgment (CJ) theory [@Thurstone_1927] is composed of two key elements: the discriminal process and the law of comparative judgment.   -->

<!-- In a CJ study evaluating text quality quality of written texts -->

@@

# References {.unnumbered}
